---
title: "Gene Expression Analysis and FDR Control"
author: "Genomics Research"
date: "`r Sys.Date()`"
output: html_document
---

Firstly we show how random values(random genes with no corellation) can affect an experiment with p_value < 0.05 when the number of experiments is large. 

```{r gene_simulation_no_corellation, message= FALSE, warning=FALSE}
set.seed(42)
m <- 20000
p_values <- runif(m, 0, 1) # Null hypothesis is true for all

# Expected false positives at alpha = 0.05
alpha <- 0.05
false_positives <- sum(p_values < alpha)
print(paste("Expected False Positives:", m * alpha))
print(paste("Actual Simulated False Positives:", false_positives))

# Visual: Distribution of P-values (Page 2)
hist(p_values, breaks = 50, col = "skyblue", main = "P-value Distribution",
     xlab = "P-value", ylab = "Frequency")
abline(v = alpha, col = "red", lwd = 2, lty = 2)
```
We can clearly see that there are some values that were tagged as significant but in reallity the correlation is non existant (random genes)


In this section we implement the Benjamin-Hochberg Procedure vs the Bonferroni

```{r param}
m <- 20 # Number of tests
alpha <- 0.05
k <- 1:m

# 1. Bonferroni Threshold (FWER)
# Constant line: alpha / m
bonf_threshold <- rep(alpha / m, m) 

# 2. BH Threshold (FDR)
# Linear line: (k / m) * alpha
bh_threshold <- (k / m) * alpha 

# Simulated P-values (some signal, some noise)
set.seed(123)
p_values <- sort(c(runif(5, 0, 0.01), runif(15, 0.05, 1)))

# Plotting the Comparison
plot(k, p_values, pch = 19, col = "black", 
     main = "FWER vs. FDR Thresholds",
     xlab = "Rank of P-value (k)", ylab = "P-value")
lines(k, bonf_threshold, col = "red", lwd = 2, lty = 2) # Bonferroni
lines(k, bh_threshold, col = "blue", lwd = 2)           # FDR/BH

legend("topleft", legend = c("P-values", "Bonferroni (FWER)", "BH (FDR)"),
       col = c("black", "red", "blue"), lty = c(NA, 2, 1), pch = c(19, NA, NA))
       
```





```{r fscd}

# Set seed so the random numbers are exactly the same every time you run it
set.seed(42)

# 1. Simulate the Data
n_genes <- 1000
n_samples_per_group <- 10
alpha <- 0.05

# We will simulate 1000 genes. 
# The first 900 have NO real difference (Null is true).
# The last 100 have a REAL difference (Alternative is true).

# Control Group (Group A)
group_A <- matrix(rnorm(n_genes * n_samples_per_group, mean = 10, sd = 2), nrow = n_genes)

# Treatment Group (Group B)
group_B_null <- matrix(rnorm(900 * n_samples_per_group, mean = 10, sd = 2), nrow = 900)
group_B_diff <- matrix(rnorm(100 * n_samples_per_group, mean = 13, sd = 2), nrow = 100) 
group_B <- rbind(group_B_null, group_B_diff)

# 2. Run t-tests on all 1000 genes
p_values <- numeric(n_genes)
for(i in 1:n_genes) {
  test <- t.test(group_A[i, ], group_B[i, ])
  p_values[i] <- test$p.value
}

# 3. Apply the Corrections
# A) Uncorrected
sig_uncorrected <- p_values < alpha

# B) FWER: Bonferroni
p_bonferroni <- p.adjust(p_values, method = "bonferroni")
sig_bonferroni <- p_bonferroni < alpha

# C) FDR: Benjamini-Hochberg (BH)
p_fdr <- p.adjust(p_values, method = "BH")
sig_fdr <- p_fdr < alpha

# 4. Tally the Results!
# Remember: Genes 1-900 are Null (False Positives if flagged)
# Genes 901-1000 are True signals (True Positives if flagged)

results <- data.frame(
  Method = c("Uncorrected (p < 0.05)", "Bonferroni (FWER)", "Benjamini-Hochberg (FDR)"),
  Total_Flags = c(sum(sig_uncorrected), sum(sig_bonferroni), sum(sig_fdr)),
  False_Positives = c(sum(sig_uncorrected[1:900]), sum(sig_bonferroni[1:900]), sum(sig_fdr[1:900])),
  True_Positives = c(sum(sig_uncorrected[901:1000]), sum(sig_bonferroni[901:1000]), sum(sig_fdr[901:1000]))
)

print("--- Case Study Results ---")
print(results)
```



```{r asdsad}
# Make sure to install ggplot2 if you haven't already: install.packages("ggplot2")
library(ggplot2)

# 1. Prepare the data for plotting
# We need to track which genes were actually true differences vs. nulls
gene_type <- c(rep("Null (Noise)", 900), rep("True Signal", 100))

plot_data <- data.frame(
  Gene_ID = 1:n_genes,
  P_Value = p_values,
  Type = gene_type
)

# 2. Sort the data by P-value (critical for the BH procedure)
plot_data <- plot_data[order(plot_data$P_Value), ]
plot_data$Rank <- 1:n_genes

# 3. Calculate the threshold lines
# Bonferroni is a flat line: alpha / m
bonferroni_threshold <- alpha / n_genes

# Benjamini-Hochberg is a diagonal line: (Rank / m) * alpha
plot_data$BH_Threshold <- (plot_data$Rank / n_genes) * alpha

# 4. Create the plot
# Note: We zoom in on the bottom left corner (lowest p-values) 
# because that is where the action happens!
p <- ggplot(plot_data, aes(x = Rank, y = P_Value)) +
  # Plot the actual p-values
  geom_point(aes(color = Type), alpha = 0.8, size = 2.5) +
  
  # Draw the Bonferroni line (Red)
  geom_hline(yintercept = bonferroni_threshold, color = "red", 
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 120, y = bonferroni_threshold + 0.002, 
           label = "Bonferroni (Strict)", color = "red", hjust = 0) +
  
  # Draw the FDR line (Blue)
  geom_line(aes(y = BH_Threshold), color = "blue", 
            linetype = "dashed", linewidth = 1) +
  annotate("text", x = 120, y = (120/n_genes)*alpha - 0.002, 
           label = "FDR (Forgiving)", color = "blue", hjust = 0) +
  
  # Zoom in on the top 150 most significant genes
  coord_cartesian(xlim = c(1, 150), ylim = c(0, 0.05)) +
  
  # Aesthetics and labels
  scale_color_manual(values = c("Null (Noise)" = "gray70", "True Signal" = "darkorange")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Why FDR Works Better for Genomics",
    subtitle = "Points below the dashed lines are declared 'Significant'",
    x = "Rank of P-Value",
    y = "P-Value",
    color = "Actual Gene Status"
  ) +
  theme(legend.position = "bottom")

# Display the plot
print(p)
```