---
title: "Gene Expression Analysis and FDR Control"
author: "Genomics Research"
date: "`r Sys.Date()`"
output: html_document
---

Firstly, we do a fake gene expression experiment to see which genes are expressed differently across two groups (e.g. cancer and control). By randomly generating 20,000 genes with randomly assigned labels, we ensure that each hypothesis is actually null (i.e. there is no real difference between the genes across groups). Running a t‑test for each gene then shows how many “significant” p‑values appear by chance when performing thousands of tests simultaneously.

```{r gene_simulation_no_corellation, message= FALSE, warning=FALSE}
set.seed(42)

# Simulate a realistic gene expression matrix
m <- 20000   # number of genes
n1 <- 50     # group 1 samples (control)
n2 <- 50     # group 2 samples (cancer)
n <- n1 + n2 # total

# Gene expression: each gene has only random noise
X <- matrix(rnorm(m * n), nrow = m)

# Create random group labels (e.g. cancer vs control)
labels <- factor(c(rep("A", n1), rep("B", n2)))

# Run t-tests for each gene using the labels
p_values <- apply(X, 1, function(gene) {
  t.test(gene ~ labels)$p.value
})

# Expected false positives at alpha = 0.05
alpha <- 0.05
false_positives <- sum(p_values < alpha)

print(paste("Expected False Positives:", m * alpha))
print(paste("Actual Simulated False Positives:", false_positives))

# Visual: Distribution of P-values
hist(p_values, breaks = 50, col = "skyblue",
     main = "P-value Distribution for Random Labels and Genes",
     xlab = "P-value", ylab = "Frequency")
abline(v = alpha, col = "red", lwd = 2, lty = 2)
```
We can clearly see that there are some values that were tagged as significant but in reality the correlation is non-existent, as the genes are purely random.


In this section we implement the Benjamin-Hochberg Procedure vs the Bonferroni

```{r FDRvsFWER}
m <- 20 # Number of tests
alpha <- 0.05
k <- 1:m

# Bonferroni Threshold (FWER)
# Constant line: alpha / m
bonf_threshold <- rep(alpha / m, m) 

# BH Threshold (FDR)
# Linear line: (k / m) * alpha
bh_threshold <- (k / m) * alpha 

# Simulated P-values (some signal, some noise)
set.seed(123)
p_values <- sort(c(runif(5, 0, 0.01), runif(15, 0.05, 1)))

# Plotting the Comparison
plot(k, p_values, pch = 19, col = "black", 
     main = "FWER vs. FDR Thresholds",
     xlab = "Rank of P-value (k)", ylab = "P-value")
lines(k, bonf_threshold, col = "red", lwd = 2, lty = 2) # Bonferroni
lines(k, bh_threshold, col = "blue", lwd = 2)           # FDR/BH

legend("topleft", legend = c("P-values", "Bonferroni (FWER)", "BH (FDR)"),
       col = c("black", "red", "blue"), lty = c(NA, 2, 1), pch = c(19, NA, NA))
       
```



In the code below we are using simulation data to make sure we are aware of the true values of false and true positives to be able to compare FDR to FWER.


```{r simulation_data_case_study}

# Set seed so the random numbers are exactly the same every time you run it
set.seed(42)

# Simulate the Data
n_genes <- 1000
n_samples_per_group <- 10
alpha <- 0.05

# We will simulate 1000 genes. 
# The first 900 have NO real difference (Null hypothesis is true).
# The last 100 have a REAL difference (Decline the null hypothesis).

# Control Group (Group A)
group_A <- matrix(rnorm(n_genes * n_samples_per_group, mean = 10, sd = 2), nrow = n_genes)

# Treatment Group (Group B)
group_B_null <- matrix(rnorm(900 * n_samples_per_group, mean = 10, sd = 2), nrow = 900)
group_B_diff <- matrix(rnorm(100 * n_samples_per_group, mean = 13, sd = 2), nrow = 100) 
group_B <- rbind(group_B_null, group_B_diff)

# Run t-tests on all 1000 genes
p_values <- numeric(n_genes)
for(i in 1:n_genes) {
  test <- t.test(group_A[i, ], group_B[i, ])
  p_values[i] <- test$p.value
}

# Apply the Corrections
# Uncorrected p_values
sig_uncorrected <- p_values < alpha

# FWER: Bonferroni
p_bonferroni <- p.adjust(p_values, method = "bonferroni")
sig_bonferroni <- p_bonferroni < alpha

# FDR: Benjamini-Hochberg (BH)
p_fdr <- p.adjust(p_values, method = "BH")
sig_fdr <- p_fdr < alpha

# Remember: Genes 1-900 are Null (False Positives if flagged)
# Genes 901-1000 are True signals (True Positives if flagged)

results <- data.frame(
  Method = c("Uncorrected (p < 0.05)", "Bonferroni (FWER)", "Benjamini-Hochberg (FDR)"),
  Total_Flags = c(sum(sig_uncorrected), sum(sig_bonferroni), sum(sig_fdr)),
  False_Positives = c(sum(sig_uncorrected[1:900]), sum(sig_bonferroni[1:900]), sum(sig_fdr[1:900])),
  True_Positives = c(sum(sig_uncorrected[901:1000]), sum(sig_bonferroni[901:1000]), sum(sig_fdr[901:1000]))
)

print(results)
```
We can certainly tell that there is a significant diference between the two methods. Bonferroni method is always correct but did only flaged 5 p_values as significant, while Benjamini-Hochberg procedure correctly pointed to 51 true positive values and 2 false positive. While 49 true positives were not flagged by FDR.


```{r plot_simulation_results}
library(ggplot2)

# We need to track which genes were actually true differences vs. nulls
gene_type <- c(rep("Null (Noise)", 900), rep("True Signal", 100))

plot_data <- data.frame(
  Gene_ID = 1:n_genes,
  P_Value = p_values,
  Type = gene_type
)

# Sort the data by P-value (critical for the BH procedure)
plot_data <- plot_data[order(plot_data$P_Value), ]
plot_data$Rank <- 1:n_genes

# Calculate the threshold lines
# Bonferroni is a flat line: alpha / m
bonferroni_threshold <- alpha / n_genes

# Benjamini-Hochberg is a diagonal line: (Rank / m) * alpha
plot_data$BH_Threshold <- (plot_data$Rank / n_genes) * alpha

# Create the plot
# Note: We zoom in on the bottom left corner (lowest p-values) because that is where p_values < 0.05 are 
p <- ggplot(plot_data, aes(x = Rank, y = P_Value)) +
  # Plot the actual p-values
  geom_point(aes(color = Type), alpha = 0.8, size = 2.5) +
  
  # Draw the Bonferroni line (Red)
  geom_hline(yintercept = bonferroni_threshold, color = "red", 
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 120, y = bonferroni_threshold + 0.002, 
           label = "Bonferroni (Strict)", color = "red", hjust = 0) +
  
  # Draw the FDR line (Blue)
  geom_line(aes(y = BH_Threshold), color = "blue", 
            linetype = "dashed", linewidth = 1) +
  annotate("text", x = 120, y = (120/n_genes)*alpha - 0.002, 
           label = "FDR (Forgiving)", color = "blue", hjust = 0) +
  
  # Zoom in on the top 150 most significant genes
  coord_cartesian(xlim = c(1, 150), ylim = c(0, 0.05)) +
  
  # Aesthetics and labels
  scale_color_manual(values = c("Null (Noise)" = "gray70", "True Signal" = "darkorange")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "FDR vs FWER for Genomics",
    subtitle = "Points below the dashed lines are declared 'Significant'",
    x = "Rank of P-Value",
    y = "P-Value",
    color = "Actual Gene Status"
  ) +
  theme(legend.position = "bottom")

# Display the plot
print(p)
```
The visualization helps us better understand the outcomes of our simulation data


Real case study example, with gene data for leukemia. It is more difficult to draw the difference between the methods here because we dont know exactly the number of true signals.


```{r real_gene_case_study, message=FALSE, warning=FALSE}
install.packages("ISLR")
library(ISLR)

# Load the Microarray Dataset
data(NCI60)
gene_data <- NCI60$data      # The actual gene expression numbers (64 rows, 6830 columns)
cancer_types <- NCI60$labs   # The labels for each of the 64 cell lines

# We want to compare Leukemia cell lines vs. all other cancer cell lines
is_leukemia <- cancer_types == "LEUKEMIA"

# Run 6,830 T-tests (One for every single gene)
# We use apply() to quickly loop through all 6830 columns 
real_p_values <- apply(gene_data, 2, function(gene_expression) {
  # Run a t-test comparing Leukemia expression vs. Other cancer expression
  test_result <- t.test(gene_expression[is_leukemia], gene_expression[!is_leukemia])
  return(test_result$p.value)
})

# Apply the Corrections
alpha <- 0.05

# Find how many are significant under each method
sig_uncorrected <- real_p_values < alpha
sig_bonferroni <- p.adjust(real_p_values, method = "bonferroni") < alpha
sig_fdr <- p.adjust(real_p_values, method = "BH") < alpha

# Summarize the Real-World Impact
real_results <- data.frame(
  Method = c("Uncorrected (p < 0.05)", "Bonferroni (FWER)", "Benjamini-Hochberg (FDR)"),
  Significant_Genes_Found = c(sum(sig_uncorrected), sum(sig_bonferroni), sum(sig_fdr))
)

cat("Total Genes Tested:", length(real_p_values), "\n\n")
print(real_results)
```
We can clearly see how much more stricter Bonferroni method is.

Gemini was used to help us produce the code for the visualization

```{r plot_real_data_ranked, fig.width=9, fig.height=6}
library(ggplot2)

# Sort the real p-values and assign ranks
sorted_real_p <- sort(real_p_values)
m_real <- length(sorted_real_p)
rank_real <- 1:m_real
alpha <- 0.05

# Calculate the thresholds for the real dataset
bonf_thresh_real <- alpha / m_real
bh_thresh_real <- (rank_real / m_real) * alpha

# Create a dataframe for ggplot
plot_data_real <- data.frame(
  Rank = rank_real,
  P_Value = sorted_real_p,
  BH_Threshold = bh_thresh_real
)

# Create the plot
ggplot(plot_data_real, aes(x = Rank, y = P_Value)) +
  # Plot the actual real gene p-values
  geom_point(color = "black", alpha = 0.5, size = 1.5) +
  
  # The strict Bonferroni line (Red)
  geom_hline(yintercept = bonf_thresh_real, color = "red", 
             linetype = "dashed", linewidth = 1) +
  
  # The adaptive FDR line (Blue)
  geom_line(aes(y = BH_Threshold), color = "blue", 
            linewidth = 1) +
  
  # ZOOM IN: The action happens in the bottom left corner!
  # We look at the top 2500 genes and zoom the y-axis to just above 0.05
  coord_cartesian(xlim = c(1, 2500), ylim = c(0, 0.06)) +
  
  # Add the Uncorrected 0.05 threshold for reference
  geom_hline(yintercept = alpha, color = "gray50", linetype = "dotted") +
  annotate("text", x = 2000, y = 0.052, label = "Uncorrected (0.05)", color = "gray50") +
  annotate("text", x = 100, y = bonf_thresh_real + 0.002, label = "Bonferroni", color = "red", hjust = 0) +
  annotate("text", x = 1500, y = (1500/m_real)*alpha - 0.002, label = "FDR Threshold", color = "blue", hjust = 0) +
  
  theme_minimal(base_size = 14) +
  labs(
    title = "The Discovery Gap in Real Leukemia Data",
    subtitle = paste("Testing", m_real, "genes. Notice the massive gap between Bonferroni and FDR."),
    x = "Rank of P-Value",
    y = "P-Value"
  )
```