---
title: "Gene Expression Analysis and FDR Control"
author: "Genomics Research"
date: "`r Sys.Date()`"
output: html_document
---

### Problem Definition

False discovery rate control is needed because modern statistical analyses often involve testing thousands of hypotheses at once, such as in genomics. Here, even a relatively low per‑test error rate leads to a flood of false positives simply because of chance. Traditional methods like the Bonferroni correction protect against any false positives but become so strict that they eliminate real discoveries. FDR offers a middle ground because it limits the expected proportion of false positives among the results declared significant. This allows us to detect meaningful differences while still keeping false discoveries under control. In the following, we will explore the FDR in practice and examine how it behaves under different testing scenarios.

### Why use the FDR?

Firstly, we do a fake gene expression experiment to see whether we can identify "genes" that are expressed differently across two groups (e.g. cancer and control). Through this, we want to demonstrate that we can make "discoveries" purely through chance. 

By randomly generating 20,000 genes with randomly assigned labels, we ensure that each hypothesis is actually null (i.e. there is no real difference between the genes across groups). Running a t‑test for each gene then shows how many “significant” p‑values appear by chance when performing thousands of tests simultaneously.

```{r gene_simulation_no_corellation, message= FALSE, warning=FALSE}
set.seed(42)

# Simulate a realistic gene expression matrix
m <- 20000   # number of genes
n1 <- 50     # group 1 samples (control)
n2 <- 50     # group 2 samples (cancer)
n <- n1 + n2 # total

# Gene expression: each gene has only random noise
X <- matrix(rnorm(m * n), nrow = m)

# Create random group labels (e.g. cancer vs control)
labels <- factor(c(rep("A", n1), rep("B", n2)))

# Run t-tests for each gene using the labels
p_values <- apply(X, 1, function(gene) {
  t.test(gene ~ labels)$p.value
})

# Expected false positives at alpha = 0.05
alpha <- 0.05
false_positives <- sum(p_values < alpha)

print(paste("Expected False Positives:", m * alpha))
print(paste("Actual Simulated False Positives:", false_positives))

# Visual: Distribution of P-values
hist(p_values, breaks = 50, col = "skyblue",
     main = "P-value Distribution for Random Labels and Genes",
     xlab = "P-value", ylab = "Frequency")
abline(v = alpha, col = "red", lwd = 2, lty = 2)
```
We can clearly see that there are some p-values that were tagged as significant but in reality the correlation is non-existent, as the genes are purely random. To be exact, there are 1042 rejected hypotheses, which is around \alpha = 5% of 20,000 and also the expected false positive rate. 

This shows that we somehow need to control for false discoveries in multiple hypothesis testing. Else, these results of these statistical tests cannot be trusted. 


### Bonferroni vs. Benjamini Hochberg

In this section, we implement the Benjamini–Hochberg procedure alongside the Bonferroni correction. Both methods set a significance cutoff to decide which hypotheses are rejected, but they do it differently. Bonferroni uses one fixed and very strict threshold for all tests and tries to prevent any false positives, while the Benjamini–Hochberg procedure assigns a separate threshold to each ranked p‑value to control the rate of false positives. This BH cutoff grows with the rank k, as it is defined as (k/m)q , so higher‑ranked p‑values are compared against a larger allowable value. Once a rank k is found where the p-value is larger than the threshold for that rank, all following p-values are not rejected.

```{r FDRvsFWER}
# --- Use the p-values from your original simulation ---
# p_values already exists from the t-tests on 20,000 genes

alpha <- 0.05
m <- length(p_values)

# BULLETPROOF FIX: Only try to show 70 if you actually have at least 70!
num_to_show <- min(70, m) 
k_show <- 1:num_to_show
p_sorted <- sort(p_values)[k_show]

# Corresponding thresholds
bonf_threshold <- rep(alpha / m, length(k_show))
bh_threshold <- (k_show / m) * alpha

# 1. Plotting the Data
plot(k_show, p_sorted,
     pch = 19, col = "black",
     main = "Bonferroni (FWER) vs BH (FDR) for Generated Gene Data",
     xlab = "Rank of P-value (k)",
     ylab = "P-value",
     ylim = c(0, max(p_sorted, na.rm = TRUE)))

lines(k_show, bonf_threshold, col = "red", lwd = 2, lty = 2)   # Bonferroni
lines(k_show, bh_threshold, col = "blue", lwd = 2)             # BH

legend("topleft",
       legend = c("P-values", "Bonferroni (FWER)", "BH (FDR)"),
       col = c("black", "red", "blue"),
       pch = c(19, NA, NA),
       lty = c(NA, 2, 1),
       lwd = c(NA, 2, 2))

# 2. Calculate the Adjusted P-values
p_adj_bonferroni <- p.adjust(p_values, method = "bonferroni")
p_adj_fdr <- p.adjust(p_values, method = "BH")

# 3. Create a clean table of the Top 10 Genes
clean_table <- data.frame(
  Rank = 1:10,
  Raw_P_Value = sort(p_values)[1:10],
  Adjusted_Bonferroni = sort(p_adj_bonferroni)[1:10],
  Adjusted_FDR = sort(p_adj_fdr)[1:10]
)

library(knitr)

# Print the Top 10 Table
kable(clean_table, 
      digits = 5,
      col.names = c("Rank", "Raw P-Value", "Bonferroni (Adjusted)", "FDR (Adjusted)"),
      caption = "Top 10 Genes: Are the adjusted values < 0.05?")

# 4. Create a final "Total Significant" Tally for 0.05
tally_table <- data.frame(
  Method = c("Bonferroni (FWER)", "Benjamini-Hochberg (FDR)"),
  Significant_at_0.05 = c(sum(p_adj_bonferroni < 0.05), sum(p_adj_fdr < 0.05))
)

# Print the Tally
kable(tally_table, caption = "Total Discoveries at Alpha = 0.05")
```

As we can see, the BH falsely rejects only 53 hypotheses compared to the 1042 that were falsely rejected before, so it successfully prevents a large rate of false discoveries. 

Bonferroni rejects even less hypotheses (only 5) and therefore performs better for this example. However, this comes at the cost of being extremely conservative. Bonferroni avoids false positives by setting a threshold so strict that it also suppresses almost all true discoveries. BH on the other hand keeps false discoveries low while still allowing a meaningful number of hypotheses to pass the threshold, making it far more useful in large‑scale testing settings where some true effects are expected and we want to be more explorative.

### How does this affect the making of discoveries?


In the following section, we utilize simulated data to establish a known 'ground truth.' By generating our own dataset where the exact number of true underlying effects is predetermined, we can precisely calculate the true positive and false positive rates. This allows for an objective, mathematical comparison between the performance of the FDR and FWER correction methods


```{r simulation_data_case_study}

# Set seed so the random numbers are exactly the same every time you run it
set.seed(42)

# Simulate the Data
n_genes <- 1000
n_samples_per_group <- 10
alpha <- 0.05

# We will simulate 1000 genes. 
# The first 900 have NO real difference (Null hypothesis is true).
# The last 100 have a REAL difference (Decline the null hypothesis).

# Control Group (Group A)
group_A <- matrix(rnorm(n_genes * n_samples_per_group, mean = 10, sd = 2), nrow = n_genes)

# Treatment Group (Group B)
group_B_null <- matrix(rnorm(900 * n_samples_per_group, mean = 10, sd = 2), nrow = 900)
group_B_diff <- matrix(rnorm(100 * n_samples_per_group, mean = 13, sd = 2), nrow = 100) 
group_B <- rbind(group_B_null, group_B_diff)

# Run t-tests on all 1000 genes
p_values <- numeric(n_genes)
for(i in 1:n_genes) {
  test <- t.test(group_A[i, ], group_B[i, ])
  p_values[i] <- test$p.value
}

# Apply the Corrections
# Uncorrected p_values
sig_uncorrected <- p_values < alpha

# FWER: Bonferroni
p_bonferroni <- p.adjust(p_values, method = "bonferroni")
sig_bonferroni <- p_bonferroni < alpha

# FDR: Benjamini-Hochberg (BH)
p_fdr <- p.adjust(p_values, method = "BH")
sig_fdr <- p_fdr < alpha

# Remember: Genes 1-900 are Null (False Positives if flagged)
# Genes 901-1000 are True signals (True Positives if flagged)

results <- data.frame(
  Method = c("Uncorrected (p < 0.05)", "Bonferroni (FWER)", "Benjamini-Hochberg (FDR)"),
  Total_Flags = c(sum(sig_uncorrected), sum(sig_bonferroni), sum(sig_fdr)),
  False_Positives = c(sum(sig_uncorrected[1:900]), sum(sig_bonferroni[1:900]), sum(sig_fdr[1:900])),
  True_Positives = c(sum(sig_uncorrected[901:1000]), sum(sig_bonferroni[901:1000]), sum(sig_fdr[901:1000]))
)

print(results)
```




```{r plot_simulation_results}
library(ggplot2)

# We need to track which genes were actually true differences vs. nulls
gene_type <- c(rep("Null (Noise)", 900), rep("True Signal", 100))

plot_data <- data.frame(
  Gene_ID = 1:n_genes,
  P_Value = p_values,
  Type = gene_type
)

# Sort the data by P-value (critical for the BH procedure)
plot_data <- plot_data[order(plot_data$P_Value), ]
plot_data$Rank <- 1:n_genes

# Calculate the threshold lines
# Bonferroni is a flat line: alpha / m
bonferroni_threshold <- alpha / n_genes

# Benjamini-Hochberg is a diagonal line: (Rank / m) * alpha
plot_data$BH_Threshold <- (plot_data$Rank / n_genes) * alpha

# Create the plot
# Note: We zoom in on the bottom left corner (lowest p-values) because that is where p_values < 0.05 are 
p <- ggplot(plot_data, aes(x = Rank, y = P_Value)) +
  # Plot the actual p-values
  geom_point(aes(color = Type), alpha = 0.8, size = 2.5) +
  
  # Draw the Bonferroni line (Red)
  geom_hline(yintercept = bonferroni_threshold, color = "red", 
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 120, y = bonferroni_threshold + 0.002, 
           label = "Bonferroni (Strict)", color = "red", hjust = 0) +
  
  # Draw the FDR line (Blue)
  geom_line(aes(y = BH_Threshold), color = "blue", 
            linetype = "dashed", linewidth = 1) +
  annotate("text", x = 120, y = (120/n_genes)*alpha - 0.002, 
           label = "FDR (Forgiving)", color = "blue", hjust = 0) +
  
  # Zoom in on the top 150 most significant genes
  coord_cartesian(xlim = c(1, 150), ylim = c(0, 0.05)) +
  
  # Aesthetics and labels
  scale_color_manual(values = c("Null (Noise)" = "gray70", "True Signal" = "darkorange")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "FDR vs FWER for Genomics",
    subtitle = "Points below the dashed lines are declared 'Significant'",
    x = "Rank of P-Value",
    y = "P-Value",
    color = "Actual Gene Status"
  ) +
  theme(legend.position = "bottom")

# Display the plot
print(p)
```
The visualization helps us better understand the outcomes of our simulation data


For our final case study, we are moving from simulation to reality using actual Leukemia gene data. This is where statistical analysis gets challenging. In our earlier simulation, we played 'creator'—we knew exactly which genes were real and which were just noise. Here in the real world, we don't have a known 'ground truth.' Because we don't know the exact number of true biological signals, we can't perfectly count our false positives. This is exactly why we rely on the mathematical framework of the False Discovery Rate that we just proved.

```{r real_gene_case_study, message=FALSE, warning=FALSE}
install.packages("ISLR")
library(ISLR)

# Load the Microarray Dataset
data(NCI60)
gene_data <- NCI60$data      # The actual gene expression numbers (64 rows, 6830 columns)
cancer_types <- NCI60$labs   # The labels for each of the 64 cell lines

# We want to compare Leukemia cell lines vs. all other cancer cell lines
is_leukemia <- cancer_types == "LEUKEMIA"

# Run 6,830 T-tests (One for every single gene)
# We use apply() to quickly loop through all 6830 columns 
real_p_values <- apply(gene_data, 2, function(gene_expression) {
  # Run a t-test comparing Leukemia expression vs. Other cancer expression
  test_result <- t.test(gene_expression[is_leukemia], gene_expression[!is_leukemia])
  return(test_result$p.value)
})

# Apply the Corrections
alpha <- 0.05

# Find how many are significant under each method
sig_uncorrected <- real_p_values < alpha
sig_bonferroni <- p.adjust(real_p_values, method = "bonferroni") < alpha
sig_fdr <- p.adjust(real_p_values, method = "BH") < alpha

# Summarize the Real-World Impact
real_results <- data.frame(
  Method = c("Uncorrected (p < 0.05)", "Bonferroni (FWER)", "Benjamini-Hochberg (FDR)"),
  Significant_Genes_Found = c(sum(sig_uncorrected), sum(sig_bonferroni), sum(sig_fdr))
)

cat("Total Genes Tested:", length(real_p_values), "\n\n")
print(real_results)
```
We can clearly see how much more stricter Bonferroni method is.

Gemini was used to help us produce the code for the visualization

```{r plot_real_data_ranked, fig.width=9, fig.height=6}
library(ggplot2)

# Sort the real p-values and assign ranks
sorted_real_p <- sort(real_p_values)
m_real <- length(sorted_real_p)
rank_real <- 1:m_real
alpha <- 0.05

# Calculate the thresholds for the real dataset
bonf_thresh_real <- alpha / m_real
bh_thresh_real <- (rank_real / m_real) * alpha

# Create a dataframe for ggplot
plot_data_real <- data.frame(
  Rank = rank_real,
  P_Value = sorted_real_p,
  BH_Threshold = bh_thresh_real
)

# Create the plot
ggplot(plot_data_real, aes(x = Rank, y = P_Value)) +
  # Plot the actual real gene p-values
  geom_point(color = "black", alpha = 0.5, size = 1.5) +
  
  # The strict Bonferroni line (Red)
  geom_hline(yintercept = bonf_thresh_real, color = "red", 
             linetype = "dashed", linewidth = 1) +
  
  # The adaptive FDR line (Blue)
  geom_line(aes(y = BH_Threshold), color = "blue", 
            linewidth = 1) +
  
  # ZOOM IN: The action happens in the bottom left corner!
  # We look at the top 2500 genes and zoom the y-axis to just above 0.05
  coord_cartesian(xlim = c(1, 2500), ylim = c(0, 0.06)) +
  
  # Add the Uncorrected 0.05 threshold for reference
  geom_hline(yintercept = alpha, color = "gray50", linetype = "dotted") +
  annotate("text", x = 2000, y = 0.052, label = "Uncorrected (0.05)", color = "gray50") +
  annotate("text", x = 100, y = bonf_thresh_real + 0.002, label = "Bonferroni", color = "red", hjust = 0) +
  annotate("text", x = 1500, y = (1500/m_real)*alpha - 0.002, label = "FDR Threshold", color = "blue", hjust = 0) +
  
  theme_minimal(base_size = 14) +
  labs(
    title = "The Discovery Gap in Real Leukemia Data",
    subtitle = paste("Testing", m_real, "genes. Notice the massive gap between Bonferroni and FDR."),
    x = "Rank of P-Value",
    y = "P-Value"
  )
```