---
title: "Gene Expression Analysis and FDR Control"
author: "Genomics Research"
date: "`r Sys.Date()`"
output: html_document
---

Firstly, we do a fake gene expression experiment to see which genes are expressed differently across two groups (e.g. cancer and control). By randomly generating 20,000 genes with randomly assigned labels, we ensure that each hypothesis is actually null (i.e. there is no real difference between the genes across groups). Running a t‑test for each gene then shows how many “significant” p‑values appear by chance when performing thousands of tests simultaneously.

```{r gene_simulation_no_corellation, message= FALSE, warning=FALSE}
set.seed(42)

# Simulate a realistic gene expression matrix
m <- 20000   # number of genes
n1 <- 50     # group 1 samples (control)
n2 <- 50     # group 2 samples (cancer)
n <- n1 + n2 # total

# Gene expression: each gene has only random noise
X <- matrix(rnorm(m * n), nrow = m)

# Create random group labels (e.g. cancer vs control)
labels <- factor(c(rep("A", n1), rep("B", n2)))

# Run t-tests for each gene using the labels
p_values <- apply(X, 1, function(gene) {
  t.test(gene ~ labels_perm)$p.value
})

# Expected false positives at alpha = 0.05
alpha <- 0.05
false_positives <- sum(p_values < alpha)

print(paste("Expected False Positives:", m * alpha))
print(paste("Actual Simulated False Positives:", false_positives))

# Visual: Distribution of P-values
hist(p_values, breaks = 50, col = "skyblue",
     main = "P-value Distribution for Random Labels and Genes",
     xlab = "P-value", ylab = "Frequency")
abline(v = alpha, col = "red", lwd = 2, lty = 2)
```
We can clearly see that there are some values that were tagged as significant but in reality the correlation is non-existent, as the genes are purely random.





In this section we implement the Benjamin-Hochberg Procedure which is a linear threshold

```{r Benjamini-Hochberg Procedure, message=FALSE, warning=FALSE}
m_small <- 10
alpha_fdr <- 0.10
p_vals_small <- c(0.001, 0.005, 0.015, 0.03, 0.045, 0.12, 0.15, 0.25, 0.50, 0.80)

# Calculate BH thresholds: (i/m) * alpha
bh_thresholds <- (1:m_small / m_small) * alpha_fdr

# Visual: BH Plot (Page 5)
plot(1:m_small, p_vals_small, pch = 19, col = "blue",
     main = "BH Procedure Threshold", xlab = "Rank (k)", ylab = "P-value")
lines(1:m_small, bh_thresholds, col = "red", lwd = 2)
legend("topleft", legend = c("P-values", "BH Threshold (alpha=0.1)"),
       col = c("blue", "red"), pch = c(19, NA), lty = c(NA, 1))
    
```


```# Mock data for 9 populations (Responders vs Non-Responders)
# Structure: Treatment A (Success, Fail), Treatment B (Success, Fail)
data_list <- list(
  pop1 = matrix(c(10, 2, 2, 10), 2), # Strong difference
  pop2 = matrix(c(8, 4, 3, 9), 2),
  pop3 = matrix(c(12, 0, 5, 7), 2),  # Strong difference
  pop4 = matrix(c(6, 6, 5, 7), 2),
  pop5 = matrix(c(7, 5, 8, 4), 2),
  pop6 = matrix(c(11, 1, 6, 6), 2),
  pop7 = matrix(c(5, 7, 6, 6), 2),
  pop8 = matrix(c(9, 3, 4, 8), 2),
  pop9 = matrix(c(10, 2, 9, 3), 2)
)

# Apply Fisher's Exact Test to each population
raw_p_values <- sapply(data_list, function(x) fisher.test(x)$p.value)

# Adjustments (Page 8-9)
bonferroni_p <- p.adjust(raw_p_values, method = "bonferroni")
fdr_p <- p.adjust(raw_p_values, method = "fdr") # Benjamini-Hochberg

# Results Comparison Table
results <- data.frame(
  Population = 1:9,
  Raw_P = round(raw_p_values, 4),
  Bonferroni = round(bonferroni_p, 4),
  FDR_BH = round(fdr_p, 4)
)
print(results)

# Count discoveries at alpha = 0.05
cat("Discoveries (Bonferroni):", sum(bonferroni_p < 0.05), "\n")
cat("Discoveries (FDR):", sum(fdr_p < 0.05), "\n")

```

```{r gene_analysis_full, message=FALSE, warning=FALSE}

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Required Bioconductor packages
packages <- c("DESeq2", "airway", "ggplot2")
for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) BiocManager::install(pkg)
}

library(DESeq2)
library(airway)
library(ggplot2)

# Load the airway dataset (RNA-seq of smooth muscle cells)
data("airway")

# Create DESeq2 object; 'dex' is the treatment (dexamethasone)
dds <- DESeqDataSet(airway, design = ~ dex)

# Filter: Remove genes with very low counts (less than 10 total reads)
dds <- dds[rowSums(counts(dds)) >= 10, ]

# The DESeq function performs normalization and statistical testing
dds <- DESeq(dds)

# results() extracts the table and applies Benjamini-Hochberg FDR by default
# alpha = 0.05 is our target False Discovery Rate threshold
res <- results(dds, alpha = 0.05)

# Convert to data frame and sort by adjusted p-value (FDR)
res_df <- as.data.frame(res)
res_df <- res_df[order(res_df$padj), ]

# Define significance for plotting
res_df$sig <- ifelse(!is.na(res_df$padj) & res_df$padj < 0.05, "FDR < 0.05", "Not Sig")

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_color_manual(values = c("FDR < 0.05" = "red", "Not Sig" = "black")) +
  theme_minimal() +
  labs(title = "Volcano Plot of Gene Expression",
       subtitle = "Red points represent genes passing the 5% False Discovery Rate",
       x = "Log2 Fold Change",
       y = "-Log10 FDR (Adjusted P-value)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue")

# Display top 10 genes with the lowest FDR
print("Top 10 Significant Genes by FDR:")
head(res_df[, c("baseMean", "log2FoldChange", "padj")], 10)
```



### What this code shows you:

1.  **The Table:** You will likely see that the "No Correction" method flags thousands of genes, while the "FDR Correction" flags significantly fewer. The difference between those two numbers represents the **False Discoveries** you avoided.
2.  **The Histograms:** * In the **Raw P-value** plot, you'll see a big spike near zero, but also a "flat" background of random noise across the whole graph. 
    * In the **FDR** plot, the adjustment pushes the non-significant genes toward 1.0, "cleaning up" your results.



### Why does this happen?
Imagine you test 20,000 genes. If you use a standard p-value of 0.05, you expect **1,000 genes** ($20,000 \times 0.05$) to look "significant" purely by random chance. FDR correction (Benjamini-Hochberg) recalculates the threshold so that only about 5% of your *final list* is expected to be false, rather than 5% of *every test you ran*.



```{r comparison_chunk, message=FALSE, warning=FALSE}
library(DESeq2)
library(airway)
library(ggplot2)
library(gridExtra) # For putting plots side-by-side

# 1. Prepare Data
data("airway")
dds <- DESeqDataSet(airway, design = ~ dex)
dds <- DESeq(dds)
res <- results(dds)
res_df <- as.data.frame(res)

# 2. Count "Significant" genes at 0.05 threshold
raw_sig_count <- sum(res_df$pvalue < 0.05, na.rm = TRUE)
fdr_sig_count <- sum(res_df$padj < 0.05, na.rm = TRUE)

# 3. Create Comparison Table
comparison <- data.frame(
  Method = c("No Correction (Raw P-value)", "With FDR Correction (BH)"),
  Significant_Genes = c(raw_sig_count, fdr_sig_count)
)
print(comparison)

# 4. Visualize the Difference
# Plot A: Distribution of Raw P-values
p1 <- ggplot(res_df, aes(x = pvalue)) +
  geom_histogram(fill = "steelblue", bins = 50) +
  geom_vline(xintercept = 0.05, color = "red", linetype = "dashed") +
  labs(title = "Raw P-values", subtitle = "Red line = 0.05 cut-off") +
  theme_minimal()

# Plot B: Distribution of FDR Adjusted P-values
p2 <- ggplot(res_df, aes(x = padj)) +
  geom_histogram(fill = "darkorange", bins = 50) +
  geom_vline(xintercept = 0.05, color = "red", linetype = "dashed") +
  labs(title = "FDR Adjusted P-values", subtitle = "Notice many genes shift to the right") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2) 
```